---
title: "R Teach-In ALT 2022: Clause linkage typology using NeighborNets"
author: "Sandra Auderset"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document: 
    latex_engine: xelatex
    keep_tex: no
header-includes:
  - |
    ```{=latex}
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
      breaksymbolleft={}, 
      showspaces = false,
      showtabs = false,
      breaklines,
      commandchars=\\\{\}
    }
    ```
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy="styler",echo=TRUE,dev = "cairo_pdf",warning = FALSE,fig.align = 'center',out.width="0.9\\textwidth")
```

# Preparation
While neighbornets can be done in R, they are still experimental and do not provide good results yet. It is thus better to prepare the data in R and then use SplitsTree4 to create and visualize the NeighborNet.

We first load the necessary packages.
```{r load packages, warning=FALSE, message=FALSE}
library(cluster)
library(phangorn)
library(stringi)
library(tidystringdist)
library(tidyverse)
```

Next, we load the data.
```{r data load}
cl <- read_csv("data/clause_linkage.csv")
glimpse(cl)
```

We create a unique label for each construction. SplitsTree cannot deal with diacritics or colons in taxon names, so we will strip all diacritics and replace colons with periods.
```{r create label}
cl <- cl %>%
  unite(Language:Label, col = "LabelNN", sep = "_", remove = FALSE, na.rm = TRUE) %>%
  mutate(LabelNN = str_replace_all(LabelNN, ":", ".")) %>%
  mutate(LabelNN = str_replace_all(LabelNN, "/", ".")) %>%
  mutate(LabelNN = str_remove_all(LabelNN, "’")) %>%
  mutate(LabelNN = stri_trans_general(LabelNN, id = "Latin-ASCII")) %>%
  mutate(across(ILL_scope:Layer, as.factor)) %>%
  arrange(Language, Label)
glimpse(cl)
```

# Creating the distance matrix and exporting the nexus file

To create a neighbornet, we first need to turn our data into a distance matrix. Since we are dealing with categorical variables of mixed data types, we use Gower distance. We use the function daisy from the cluster package to calculate it.
```{r create dist}
# create distance matrix
cl_variables <- cl %>%
  select(ILL_scope:Layer)
cl_dist <- daisy(cl_variables, metric = "gower")
# add labels back
cl_dist <- as.matrix(cl_dist, labels=TRUE)
rownames(cl_dist) <- cl$LabelNN
```

Now we are ready to export the distance matrix for SplitsTree. We use the package phangorn to write it into the required nexus format.
```{r export nexus}
# export nexus file for splitstree
write.nexus.dist(cl_dist, file = "data/clause_linkage_dist.nex")
```

