---
title: "Data wrangling: WALS"
author: "Olena Shcherbakova"
date: '2022-11-07'
output:
  html_document: default
  word_document: default
  pdf_document: default
---

Before starting, we need to (install and) load necessary packages.

```{r}
library(tidyverse)
library(janitor)
```

## How were WALS-related files downloaded?

All files on WALS in cldf format are stored on Zenodo: <https://zenodo.org/record/3731125#.Y2jZOuTMJZU>. These can be downloaded and extracted on the local machine or accessed online.

```{r}
wals <- read_delim("data/cldf_wals/languages.csv") %>% #obtaining languages
  select("ID", "Name", "Glottocode") %>%
  rename(Language_ID = "ID") %>% #renaming the column to avoid problems
  left_join(read_delim(
    "data/cldf_wals/values.csv") %>% 
      select("Language_ID", "Parameter_ID", "Value")
    ) %>% #obtaining values per feature and language
  left_join(read_delim(
    "data/cldf_wals/codes.csv") %>% 
      rename(Parameter_name = "Name") %>% 
      select("Parameter_ID", "Parameter_name", 
                    "Description"))
wals
```

Online example:

```{r}
wals <- read_delim("https://raw.githubusercontent.com/cldf-datasets/wals/master/cldf/languages.csv") %>% #obtainng languages
  select("ID", "Name", "Glottocode") %>%
  rename(Language_ID = "ID") %>% #renaming the column to avoid problems
  left_join(read_delim(
    "https://raw.githubusercontent.com/cldf-datasets/wals/master/cldf/values.csv") %>%
      select("Language_ID", 
             "Parameter_ID", 
              "Value")) %>% #obtaining values per feature & language
  left_join(
    read_delim("https://raw.githubusercontent.com/cldf-datasets/wals/master/cldf/codes.csv") %>% 
      rename(Parameter_name = "Name") %>% 
      select("Parameter_ID", "Parameter_name", 
                    "Description"))
```

## Goals

-   **summarizing information on two WALS features**
-   long --\> wide dataframe format
-   wide --\> long dataframe format
-   `select`ing variables
-   `filter`ing observations (rows)
-   creating new variables

1.  we transform our dataframe from long to wide format.

```{r}
wals_wide <- wals %>%
  select(!c(Description, Parameter_name, Glottocode, Language_ID)) %>%
  distinct() %>% #removes duplicate rows based on all columns
  pivot_wider(names_from = Parameter_ID, values_from = Value)
wals_wide
```

2.  we transform dataframe from wide to long format.

```{r}
wals_long <- wals_wide %>% 
  pivot_longer(cols = !Name, #alternative: cols = `81A`:`90G`
               names_to = "Parameter_ID", values_to = "Value") 
wals_long
```

3.  we summarize the information on the order of object and verb (83A) and adposition and noun phrase (85A) from WALS by:

-   `filter`ing the rows, so that only observations with `83A` and `85A` values in `Parameter_ID` column are kept

-   `select`ing only relevant columns

-   removing duplicate rows with `distinct`

-   reshaping the dataframe into the "wide" format using `pivot_wider`

-   removing observations with missing values for both WALS features using `is.na` within `filter`

-   creating a new variable using `case_when` within `mutate`

-   removing all observations that do not fall into the 4 major groups by `filter`ing

-   count how many languages are in each remaining group using `tabyl`

Here is our `wals` dataframe, and it has a long format:

```{r}
wals
```

```{r}
wals_word_order <- wals %>%
  filter(Parameter_ID == "83A" | Parameter_ID == "85A") %>%
  select(!c(Description, Parameter_name)) %>%
  distinct() %>% #removes duplicate rows based on all columns
  pivot_wider(names_from = Parameter_ID, values_from = Value) %>%
  filter(!is.na(`83A`), !is.na(`85A`)) %>%
  mutate(Word_order_combination = 
           case_when(
    `83A` == "2" & `85A` == "1" ~ "VO + Postpositions",
    `83A` == "2" & `85A` == "2" ~ "VO + Prepositions",
    `83A` == "1" & `85A` == "1" ~ "OV + Postpositions",
    `83A` == "1" & `85A` == "2" ~ "OV + Prepositions",
    TRUE ~ "something else")
    ) %>%
  filter(Word_order_combination != "something else") %>% #removing observations with "something else" value in Word_order_combination
  tabyl(`Word_order_combination`) #counting the number of observations in each group

wals_word_order
```
